Constructing a universe for the setoid model

### 2020-05-05 paper.tex
Add paper.tex

\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\J}{\mathrm{J}}
\newcommand{\El}{\textsf{El}}

% \newcommand{\Set}{\mathbf{Set}}
% \newcommand{\Prop}{\mathbf{Prop}}
\newcommand{\Type}{\mathbf{Type}}
\renewcommand{\iff}{\leftrightarrow}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\Int}{\mathbb{Z}}
\newcommand{\Rat}{\mathbb{Q}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Bool}{\mathrm{Bool}}

@@

% \newcommand{\Set}{\mathbf{Set}}
% \newcommand{\Prop}{\mathbf{Prop}}
% \newcommand{\Type}{\mathbf{Type}}
% \renewcommand{\iff}{\leftrightarrow}
% \newcommand{\Nat}{\mathbb{N}}
% \newtheorem{theorem}{Theorem}
% \newcommand{\cont}[2]{#1 \lhd #2}
% \def\lv{\mathopen{{[\kern-0.14em[}}}    % opening [[ value delimiter
% \def\rv{\mathclose{{]\kern-0.14em]}}}   % closing ]] value delimiter
% \newcommand{\eval}[1]{{\lv{#1}\rv}}

@@

The setoid interpretation gives rise to a model of a type theory with functional
and propositional extensionality. It is thus a way to explain extensionality in
a type-theoretic and computationally adequate way \cite{setoid-model}. The
setoid model has been recently adapted into Setoid Type Theory (SeTT), which is
justified by a syntactic translation into a very basic \emph{target theory}
\cite{mpc19}. This translation relies on the existence in the target theory of a
universe of definitionally proof-irrelevant propositions $\Prop$, as recently
implemented in Coq and Agda \cite{gilbert}. The setoid interpretation models a
universe of propositions where equality of propositions is logical equivalence,
thus providing a very basic instance of univalence. However, we also would like
to be able to equip SeTT with a universe of setoids. This universe must in turn
be a setoid, hence in particular it cannot be univalent, but rather its
equivalence relation must reflect simple equality of codes.

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

However, we don't want to assume inductive-recursive types in the basic type
theory which is the target of the setoid translation. We know that we can translate
basic instances of induction-recursion into inductive families using the
equivalence of $I$-indexed families of sets and sets over $I$,
i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
\Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
  this seems to be a non-conservative extension.}:
%
\begin{align*}
  & \textsf{data}\ \U\ : \Set_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
types in $\Set$ that are in the image of $\El$:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in
$\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
universe. Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
However, we now need an inductive-inductive type in the target theory:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}

>>>>

\newcommand{\refl}{\textsf{refl}}
\newcommand{\GG}{\Gamma}
\newcommand{\DD}{\Delta}
\newcommand{\GD}{\Delta}
\newcommand{\R}{\textsf{R}}
\newcommand{\Id}{\textsf{Id}}
\newcommand{\coe}{\textsf{coe}}
\newcommand{\coh}{\textsf{coh}}
\newcommand{\J}{\textsf{J}}
\newcommand{\tyj}[2]{#2 : \Ty\ #1}
\newcommand{\tmj}[3]{#2 : \Tm\ #1\ #3}
\newcommand{\subj}[3]{#1 : \mathrm{Sub}\ #2\ #3}
\newcommand{\setoidU}{\mathcal{U}}
\newcommand{\ad}[1]{\AgdaFunction{#1}}
\newcommand{\Set}{\textsf{Set}}
\newcommand{\Prop}{\textsf{Prop}}
\newcommand{\Setoid}{\textsf{Setoid}}
\newcommand{\Con}{\mathrm{Con}}
\newcommand{\Ty}{\mathrm{Ty}}
\newcommand{\Tm}{\mathrm{Tm}}
\newcommand{\Tms}{\mathrm{Tms}}
\newcommand{\id}{\mathrm{id}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\cat}[1]{\underline{\mathbf{#1}}}
\newcommand{\U}{\textsf{U}}
\newcommand{\El}{\textsf{El}}

\newcommand{\Type}{\mathbf{Type}}
\renewcommand{\iff}{\leftrightarrow}
\newcommand{\Nat}{\mathbb{N}}
\newcommand{\Int}{\mathbb{Z}}
\newcommand{\Rat}{\mathbb{Q}}
\newcommand{\Real}{\mathbb{R}}
\newcommand{\Bool}{\mathrm{Bool}}

@@

@@

\section{Introduction}

...

Intensional Type Theory has nice computational properties, like decidability of
type-checking, that can make it more suitable for computer formalization, since
proof-checking can be implemented as an entirely automatic procedure.
%
However, a serious drawback of ITT is given by the lack of extensionality.
%
Two extensionality principles are particularly missed from intensional type
theory, in that they arguably constitute the bare minimum for type theory to be
comparable to set theory as a foundational system for set-level mathematics.
%
The first of these is functional extensionality, i.e. the principle according to
which two functions can be considered equal if they map equal inputs to equal
outputs.
%
The second principle is propositional extensionality, according to which two
propositions should be considered equal when logically equivalent.

These extensionality principles are not theorems of intensional type theory as
usually defined [cite], although they are consistent with it [cite].

As a consequence, these principles are often added as axioms when formalizing
mathematics using proof-assistants based on ITT. This solution, however, is far
from optimal, as the introduction of axioms destroys the nice computational
properties of the system, particularly canonicity. Unfortunately, one can show
that whenever equality is defined according to the usual inductive definition by
ML, there is no hope to have a theory where both canonicity and funext/propext
coexist.

The first step is therefore to give up the idea of defining equality as a single
inductive definition, and adopt the approach according to which equality should
be defined in a specific way for each type former in the type theory. That is,
every type former should come equipped with its own notion of equality, from
which a general notion of equality is derived.

Two systems of ITT with extensional properties that follow this approach are OTT
and cubical TT. In the first one ... TODO. In cubical TT ... TODO. Note that we
do not consider HoTT since, although it is a version of ITT with extensional
properties, it's not known to enjoy canonicity.

The system that we will consider in this paper follows the same approach of OTT,
namely the setoid model... TODO

* Internal encoding of setoids

* Setoid hell

* Setoid TT: MLTT + setoid constructs

* universe of setoids

\section{Setoid Type Theory}

\subsection{Setoid model}

The setoid interpretation gives rise to a model of a type theory with functional
and propositional extensionality. It is thus a way to explain extensionality in
a type-theoretic and computationally adequate way \cite{setoid-model}.

\todo{say something more about this}

\subsection{SeTT}

The setoid model has been recently adapted into Setoid Type Theory (SeTT), which
is justified by a syntactic translation into a very basic \emph{target theory}
\cite{mpc19}. This translation relies on the existence in the target theory of a
universe of definitionally proof-irrelevant propositions $\Prop$, as recently
implemented in Coq and Agda \cite{gilbert}. The setoid interpretation models a
universe of propositions where equality of propositions is logical equivalence,
thus providing a very basic instance of univalence.

Technically speaking, SeTT is an extension of MLTT with extra component to
support setoid reasoning. These extra components are nothing more than an
internalization of the idea that every type is a setoid. More precisely, we have
rules [explaining TODO] that contexts (aka closed types) are setoids, and that
dependent types are setoids indexed over setoids.

\[
\infer{
  \GG^\sim\ \rho_0\ \rho_1 : \Tm\ \DD\ \Prop
}{
  \GG : \Con \qquad \subj{\rho_0, \rho_1}{\DD}{\GG}
}
\qquad
\infer{
  A^{\sim}\ p\ a_0\ a_1 : \Tm\ \DD\ \Prop
}{
  \deduce{
    a_0 : \Tm\ \DD\ A[\rho_0] \qquad a_1 : \Tm\ \DD\ A[\rho_1]
  }{
    A : \Ty\ \GG \qquad p : \Tm\ \DD\ (\El\ (\GG^{\sim}\ \rho_0\ \rho_1))
  }
}
\]

Note that the setoid equivalence relations are strict propositions. In addition,
we also have rules representing the fact that every construction in SeTT
respects setoid equality, so that we can transport along any such equality:

[coe and coh rules here]

    \[
    \infer{
      \deduce{
        \tmj{\GD}{\coh_A\ p\ a}{\underline{A^\sim\ p\ a\ (\coe_A\ p\ a)}}
      }{
        \tmj{\GD}{\coe_A\ p\ a}{A[\rho_1]}
      }
    }{
      \tyj{\GG}{A} \qquad \subj{\rho_0, \rho_1}{\GD}{\GG} \qquad
      \tmj{\GD}{p}{\underline{\GG^\sim\ \rho_0\ \rho_1}}\qquad
      \tmj{\GD}{a}{A[\rho_0]}
    }
    \]

Equality in SeTT differs from the standard approach by Martin-Loef in that every
type comes equipped with an equality relation. This is in contrast with ML's
identity type which is defined uniformly for any type.

Nevertheless, we can easily recover the usual ML identity type from setoid
equality:

\[
\infer{
  \textsf{Id}_A\ a_0\ a_1 :\equiv A^{\sim}\ (\R\ \GG)\ a_0\ a_1 : \Tm\ \GG\ \Prop
}{
  \tyj{\GG}{A} \qquad \tmj{\GG}{a_0, a_1}{A}
}
\]

Note that ... we have UIP... Moreover, contractibility of singletons holds by
UIP and by definition of equality on $\Sigma$ types. Therefore we can define the
$\J$ eliminator:

\[
\infer{
  \tmj{\GG}{\J_M\ p\ d}{M[a_0, a_1, p]}
}{
  \deduce{
    \tmj{\GG}{d}{M[a_0, a_0, \refl_A\ a_0]}
    \qquad
    \tmj{\GG}{p}{\underline{\Id_A\ a_0\ a_1}}
  }{
  \tyj{(\GG, x : A, y : A, p : \Id_A\ x\ y)}{M}
  }
}
\]

\section{Universe of setoids}

As currently defined, SeTT is a system already powerful enough to support the
formalization of part of set-level mathematics. However, we also would like to
be able to equip SeTT with a universe of setoids. This universe must in turn be
a setoid, hence in particular it cannot be univalent, but rather its equivalence
relation must reflect simple equality of codes.

As we use Type Theory as our metatheory, this universe of setoids is going to be
an internal construction in Type Theory. The usual way to encode universes in
Type Theory is by an inductive-recursive definition. As we will see in Section
TODO, it is indeed possible to give a IR definition of a universe of setoids.
%
However, we would prefer not to be forced to assume inductive-recursive types,
or IR in general, in the basic type theory which is the target of the setoid
translation. This is because the point of the translation is to explain a
complex system in terms of a very simple, well understood core, and we think
introducing IR would go against this goal.

Fortunately, it is possible to reduce some instances of induction-recursion to
plain induction (more specifically, inductive families), and the well known IR
universe construction is indeed one of them.

In the next section, we recall the IR construction of universes in type theory,
as well as lay out the process by which such construction can be turned to one
that only relies on inductive families. In Section TODO, we give a first IR
definition of the universe of setoids. After that we will see if and how such IR
definition can be turned into one that does not rely on IR.

\subsection{Encoding of universes via IR and II}

A standard, well-known way of encoding universes in type theory is via induction-recursion.
\todo{a couple of words about induction recursion}.

To define a universe, we simultaneously define the universe of codes itself, as
an inductive definition, and a family $\El : \U \to \Set$ that assigns, to each
code corresponding to some type, the meta-theoretic type of its elements.

For example, the definition of a small universe with booleans and $\Pi$ types
would be the following:

\begin{align*}
  & \textsf{data}\ \U\ : \Set_1 \\
  & \quad \textsf{bool} : \U \\
  & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
  & \El\ \textsf{bool} = \mathbbm{2} \\
  & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
\end{align*}

Note that $\U$ is defined to be in $\Set_1$ rather than $\Set$. This is
because... \todo{TODO}

As it turns out, induction-recursion is not really required to encode such
a universe, although is certainly makes the definition particularly nice.
%
In fact, we know that we can translate basic instances of induction-recursion
into inductive families using the equivalence of $I$-indexed families of sets
and sets over $I$, i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$.

Following this, we seen that we can actually encode the universe above as an
inductive type $\ad{in-U}$ that \emph{carves out} all types in $\Set$ that are
in the image of $\El$:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
  & \quad \textsf{inPi}
  :  \{A : \Set\} \{B : A \to \Set\}\\
  & \ \qquad \to \ad{in-U}\ A
  \to ((a : A) \to \ad{in-U}\ (B\ a))
  \to \ad{in-U}\ ((a : A) \to (B\ a))
\end{align*}
Using this $\U$ and $\El$ can be given as follows:
\begin{align*}
  \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
  \El & = \textsf{proj}_1
\end{align*}

Note that this construction gives rise to a universe in $\Set_1$, rather than
$\Set$, since the definition of $\U$ quantifies over all possible sets. Thus,
the meta-theory must contain at least one universe for this construction to go
through.

%% ----------------------

%% However, we don't want to assume inductive-recursive types in the basic type
%% theory which is the target of the setoid translation. We know that we can translate
%% basic instances of induction-recursion into inductive families using the
%% equivalence of $I$-indexed families of sets and sets over $I$,
%% i.e.\ $[I, \Set_i] \simeq \Set_i/I$ for $I : \Set_i$. For example, the
%% inductive-recursive definition of a universe $\U:\Set_1$ and $\El : \U \to
%% \Set$ with $\Pi$-types and Booleans is\footnote{In Agda, this definition would also go through if $\U$ were in $\Set$, but
%%   this seems to be a non-conservative extension.}:
%% %
%% \begin{align*}
%%   & \textsf{data}\ \U\ : \Set_1 \\
%%   & \quad \textsf{bool} : \U \\
%%   & \quad \textsf{pi} : (A : \U) \to (\El\ A \to \U) \to \U \\
%%   & \El\ \textsf{bool} = \mathbbm{2} \\
%%   & \El\ (\textsf{pi}\ A\ B) = (a : \El\ A) \to \El\ (B\ a)
%% \end{align*}

%% We can model this as an inductive type $\ad{in-U}$ that \emph{carves out} all
%% types in $\Set$ that are in the image of $\El$:
%% %
%% \begin{align*}
%%   & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
%%   & \quad \textsf{inBool} : \textsf{in-U}\ \mathbbm{2} \\
%%   & \quad \textsf{inPi}
%%   :  \{A : \Set\} \{B : A \to \Set\}\\
%%   & \ \qquad \to \ad{in-U}\ A
%%   \to ((a : A) \to \ad{in-U}\ (B\ a))
%%   \to \ad{in-U}\ ((a : A) \to (B\ a))
%% \end{align*}
%% Using this $\U$ and $\El$ can be given as follows:
%% \begin{align*}
%%   \U & = \Sigma (A : \Set)\ (\textsf{in-U}\ A) \\
%%   \El & = \textsf{proj}_1
%% \end{align*}

%% Note that this construction gives rise to a universe in
%% $\Set_1$, rather than $\Set$, thus the meta-theory must contain at least one
%% universe.

\subsection{Inductive-inductive universe}

To provide such a universe we need to define a setoid $\setoidU : \Setoid$ and a
family of setoids $\El : \setoidU \to \Setoid$, with codes for basic type
formers like $\Pi$-types and booleans. This can be obtained as an inductive-recursive type.
\todo{expand definition to include more details}
%
\begin{align*}
  & \textsf{data} \ \setoidU : \Set_1 \\
  & \_\sim_\setoidU\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \El : \setoidU \to \Set \\
  & \_\vdash\_\sim_\El\_ : \{a\ a' : \setoidU\} \to a \sim_\setoidU a' \to \El\ a \to \El\ a' \to \Prop
\end{align*}

As said before, however, we would like to avoid having to assume
inductive-recursive types, of more generally induction-recursion, in the
metatheory.

Our result is that a modified form of this translation also works for
the more complex inductive-recursive type we need to model the universe of
setoids. 
%
In particular, in addition to $\ad{in-U}$ for defining $\U$ as before, we also
introduce a family $\ad{in-U∼}$ of binary relations between types in the
universe, from which we then define $\_\sim_{\setoidU}\_$.
%
However, we now need an inductive-inductive type in the target theory:
%
\begin{align*}
  & \textsf{data}\ \ad{in-U} : \Set \to \Set_1 \\
  & \textsf{data}\ \ad{in-U∼} : \{A\ A' : \Set\} \to \ad{in-U}\ A \to \ad{in-U}\ A' \to (A \to A' \to \Prop) \to \Set_1 \\
  & \setoidU : \Set_1 \\
  & \El : \setoidU \to \Set \\
  & \_\sim_{\setoidU}\_ : \setoidU \to \setoidU \to \Prop_1 \\
  & \_\vdash\_\sim_{\El}\_ : \{a\ a' : \U\} \to a\ \sim_{\setoidU}\ a' \to \El\ a \to \El\ a' \to \Prop \\
  & \\
  & \setoidU = \Sigma\ (X : \Set)\ (\ad{in-U}\ X) \\
  & \El = \textsf{proj}_1 \\
  & (X , p) \sim_{\setoidU} (X' , p') =
   \ \parallel \Sigma\ (R : X \to X' \to \Prop)\ (\ad{in-U∼}\ p\ p'\ R) \parallel
\end{align*}
